preregistration

### 2022-02-23 preregistrations/2_study1/preregistration.Rmd
changes to flanker instructions

In line with previous theorizing [e.g., @ellis_2020; @frankenhuis_2016; @frankenhuis_2013], we initially expected people from adversity to perform better on cognitive tasks that require a broad attention scope in order to successfully detect peripheral and subtly changing information, and to perform worse on tasks requiring ignoring interfering information (see the [preregistration](https://github.com/StefanVermeent/attention_project/blob/Preregistration/preregistrations/1_pilot/preregistration.docx) of the first pilot study). However, the data pointed in the opposite direction. A Drift Diffusion analysis showed that participants with more violence exposure were slower to process information and orient their attention across all three tasks. At the same time, they were faster to narrow down their attention to the central target on the Flanker Task, thereby effectively limiting the amount of flanker interference that they were exposed to.

@@

Figure 1 shows two possible data patterns that we might expect across the different versions at different levels of violence exposure. First, if the findings reflect a general information processing deficit, we would expect performance to be lower for people with more violence exposure across all versions of the task. Alternatively, lower performance on the standard version might reflect an adaptive trade-off towards more robust cognitive functioning, even in the face of noise or perturbations [@giudice_2018]. In that case, we would expect more stable performance for people with more violence exposure. As a result, they might not benefit as much from enhanced visual information, but at the same time might be able to better maintain performance when presented with degraded information.

```{r figure1, fig.width=6, dpi=600, fig.id = "figure1", fig.cap.style = "Image Caption", fig.cap="Overview of the three cognitive tasks."}
knitr::include_graphics("figure1.png", error = F)
```

## Primary aims

@@

# Secondary aims

@@

We conducted a power simulation using the *faux* package in R [@debruine_2021] to determine the optimal number of participants to include (more information including all simulation code on [Github](https://github.com/StefanVermeent/attention_project/tree/preregistration1/preregistrations/1_pilot/scripts)).

@@

We applied several exclusion criteria prior to analyzing the data. First, we excluded participants who did not complete the full study and those who did not complete all three Flanker conditions (*N* = `r #TBD `). Second, we analyzed responses to the attention checks and reversed coded items in the questionnaire part of the experiment. We excluded participants if they missed both attention check items or if they had suspicious response patterns (e.g., consistently endorsing high response options even when some items were reverse coded) (*N* = `r #TBD `).

@@

Based on participant feedback in the pilot study, we changed (the wording of) a number of questionnaire items in order to make them more inclusive. We added an item asking about the participant's family composition. In addition, all mentions of 'parent(s)' across questionnaires were changed to 'caregiver(s)' (e.g., "At least one of my caregivers had punishments that were unpredictable"). We also added a question asking about the exact family composition.

@@

After completing the attention tasks, participants completed the questionnaire battery in the following fixed order: 1) Current state (state anxiety and separate questions relating to specific states); 2) Childhood adversity (perceived unpredictability, perceived socio-economic status, exposure to violence and physical fights, household chaos); 3) Temporal orientation (impulsivity, future orientation); 4) Depressive symptoms.

Finally, the demographics questions asked about the participant's age, weight, height, physical activity, sex at birth, gender, ethnicity, social class (current and during childhood), education level (one's own as well as parents' education level), occupation, and household income (current and during childhood). At the end of the experiment, we asked participants if they ever got up or were interrupted during the study, and how noisy their environment was during the attention tasks.

@@

First, we pooled the Flanker data of the pilot study and the current study, using the standard condition of the current study. For the effect of violence exposure on each of the DDM parameters, we ran initial linear models including violence exposure, study number (pilot study or current study, sum-coded) and their interaction as an independent variable to check whether the two studies were sufficiently similar. If the interaction was non-significant, we ran a final model version including only violence exposure to estimate the main effect. For the analyses using raw RTs as the dependent variable, these analyses also included congruency as an independent variable, as well as the full three-way interaction. Similarly, we estimated a final model without study number if the interaction was non-significant.

Second, we analyzed the effect of violence exposure and Flanker condition type (within the current study only) on Flanker performance using linear mixed effects models with a random intercept per subject. The main dependent variables were mean RTs and the SSP parameters: Perceptual input (*p*), boundary separation (*a*), non-decision time (*t0*), initial attention width (*sd~a~*), attention shrinking rate (*rd*) and the interference effect (*SD~a~*/*rd*). For each outcome measure, we ran two separate models: One comparing the standard condition with the enhanced condition, and one comparing the standard condition with the degraded condition. In both models, condition was dummy-coded using the standard condition as the reference group.

>>>>

In line with previous theorizing [e.g., @ellis_2020; @frankenhuis_2016; @frankenhuis_2013], we initially expected people from adversity to perform better on cognitive tasks that require a broad attention scope in order to successfully detect peripheral and subtly changing information, and to perform worse on tasks requiring ignoring interfering information (see the [preregistration](https://github.com/StefanVermeent/attention_project/blob/Preregistration/preregistrations/1_pilot/preregistration.docx) of the first pilot study). However, the data pointed in the opposite direction. A Drift Diffusion analysis showed that participants with more violence exposure were generally slower to process information and orient their attention across all three tasks. At the same time, they were faster to narrow down their attention to the central target on the Flanker Task, thereby effectively limiting the amount of flanker interference that they were exposed to.

@@

Figure \@ref(fig:figure1) shows two possible data patterns that we might expect across the different versions at different levels of violence exposure. First, if the findings reflect a general information processing deficit, we would expect performance to be lower for people with more violence exposure across all versions of the task. Alternatively, lower performance on the standard version might reflect an adaptive trade-off towards more robust cognitive functioning, even in the face of noise or perturbations [@giudice_2018]. In that case, we would expect more stable performance for people with more violence exposure. As a result, they might not benefit as much from enhanced visual information, but at the same time might be able to better maintain performance when presented with degraded information.

```{r figure1, fig.width=6, dpi=600, fig.id = "figure1", fig.cap.style = "Image Caption", fig.cap="Overview of the three Flanker conditions and predicted patterns of results under different cognitive mechanisms. A: The different Flanker Task conditions. The standard version is a typical implementation of the Flanker stimuli. The enhanced version improves the quality of information by increasing stimulus size by 12.5% and including padding (5 pixels) between the arrows. The degraded cognition reduces the quality of information by rotating all arrows at a 45° angle. B: Two potential patterns of the effect of condition on people’s perceptual input as a function of violence exposure. Left plot: Under the deficit model, we would expect people with more violence exposure to perform worse across all three conditions. Right plot: If performance is traded off against higher robustness of the cognitive system for people with more violence exposure (Del Giudice & Crespi, 2018), we would expect cognitive performance in this group to be relatively stable across conditions, perhaps even outperforming people with less violence exposure in the degraded condition."}
knitr::include_graphics("Fig1.png", error = F)
```

## Primary aims {#primary}

@@

# Secondary  {#secondary}

@@

We conducted a power simulation using the *faux* package in R [@debruine_2021] to determine the optimal number of participants to include (more information including all simulation code on [Github](https://github.com/StefanVermeent/attention_project/tree/preregistration1/preregistrations/1_pilot/scripts)). Assuming an effect size of $\beta$ = 0.1 and a DDM parameter recovery accuracy of *r* = .85, we estimated that we would need 500 participants to achieve at least 80% power. We sampled a total of 550 participants to account for necessary exclusions (based on our experiences in the pilot study).

@@

We applied several exclusion criteria prior to analyzing the data. First, we excluded participants who did not complete the full study and those who did not complete all three Flanker conditions (*N* = `r #TBD `). Second, we analyzed responses to the attention checks and reversed coded items in the questionnaire part of the experiment. We excluded participants if they missed both attention check items or if they had suspicious response patterns (e.g., consistently endorsing high response options even when some items were reverse coded) (*N* = `r #TBD `). Third, we excluded participants whose screen height was < 700 pixels, or whose screen height was bigger than their screen width (suggesting that they did not complete the experiment on a laptop or desktop pc). Finally, we screened participants' accuracy on the Flanker task within each condition. Using a binomial distribution, we calculated that if a participant would be purely guessing on each trial, they would have a probability of 97.5% of obtaining an overall accuracy of 59.4%. Therefore, we excluded participant's data on a particular condition if their accuracy on that condition was below 60%.

@@

Based on participant feedback in the pilot study, we changed (the wording of) a number of questionnaire items in order to make them more inclusive. We added an item asking about the participant's family composition. In addition, all mentions of 'parent(s)' across questionnaires were changed to 'caregiver(s)' (e.g., "At least one of my caregivers had punishments that were unpredictable"). We also added a question asking about the participant's exact family composition.

@@

After completing the Flanker task, participants completed the questionnaire battery in the following fixed order: 1) Current state (state anxiety and separate questions relating to specific states); 2) Childhood adversity (perceived unpredictability, perceived socio-economic status, exposure to violence and physical fights, household chaos); 3) Temporal orientation (impulsivity, future orientation); 4) Depressive symptoms.

Finally, the demographics questions asked about the participant's age, weight, height, physical activity, sex at birth, gender, ethnicity, social class (current and during childhood), education level (one's own as well as caregivers' education level), occupation, and household income (current and during childhood). At the end of the experiment, we asked participants if they ever got up or were interrupted during the study, and how noisy their environment was during the attention tasks.

@@

To address the [first primary aim](#primary) we pooled the Flanker data of the pilot study and the current study, using the standard condition of the current study. For the pooled effect of violence exposure on each of the DDM parameters, we ran initial linear models including violence exposure, study number (pilot study vs. current study, sum-coded) and their interaction as an independent variable to check whether the two studies were sufficiently similar. If the interaction was non-significant, we ran a final model version including only violence exposure to estimate the main effect. The pooled effect of violence exposure on raw RTs was assessed through linear mixed models including a random intercept per participant. Initial models included violence exposure, congruency (congruent vs. incongruent, sum-coded as -1 and 1), the violence exposure X congruency interaction, study number (pilot study vs current study, sum-coded), and the violence_exposure X study number interaction. We estimated a final model without study number if the violence exposure X study number interaction was non-significant.

To address the [second primary aim](#primary), we analyzed the effect of violence exposure and Flanker condition type (within the current study only) on Flanker performance using linear mixed effects models with a random intercept per participant. The main dependent variables were mean RTs and the SSP parameters: Perceptual input (*p*), boundary separation (*a*), non-decision time (*t0*), initial attention width (*sd~a~*), attention shrinking rate (*rd*) and the interference effect (*SD~a~*/*rd*). For each outcome measure, we ran two separate models: One comparing the standard condition with the enhanced condition, and one comparing the standard condition with the degraded condition. In both models, condition was dummy-coded using the standard condition as the reference group.
